preference

### 2020-07-13 analyses/BB_abund_GAM.Rmd
update repo

### FL and BB abundance / diversity
```{r echo=FALSE, include=FALSE}
### Per-transect BB abundance: species-level
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  group_by(site, date, bb.sp) %>% # group by bb.sp*transect (and other variables we want to retain)
  summarize(bb.abund = n()) %>% # get bb.sp*transect abundance
  ungroup %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) %>% # complete implicit absences
  mutate(year = factor(year(date))) %>%
  left_join(site_data) %>%
  mutate(yday = yday(date),
         bb.sp = factor(bb.sp),
         management = factor(management)) %>%
  semi_join(net, by = c("site", "date")) %>%
  group_by(year, bb.sp) %>%
  mutate(bb.total.abund = sum(bb.abund))
```

GAM modeling of abundance per bb species
```{r}
### Call models
bb_abund.2010 <- bam(bb.abund ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(bb_abund_sp, year == 2010 & bb.total.abund >= 5),
                        discrete = TRUE,
                        family = "poisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(bb_abund.2010)
bb_abund.2010.sum <- summary(bb_abund.2010)
print(plot(bb_abund.2010, allTerms = TRUE), pages = 4)

bb_abund.2011 <- bam(bb.abund ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(bb_abund_sp, year == 2011 & bb.total.abund >= 5),
                        discrete = TRUE,
                        family = "poisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(bb_abund.2011)
bb_abund.2011.sum <- summary(bb_abund.2011)
print(plot(bb_abund.2011 , allTerms = TRUE), pages = 4)

bb_abund.2012 <- bam(bb.abund ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(bb_abund_sp, year == 2012 & bb.total.abund >= 5),
                        discrete = TRUE,
                        family = "poisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(bb_abund.2012)
bb_abund.2012.sum <- summary(bb_abund.2012)
print(plot(bb_abund.2012 , allTerms = TRUE), pages = 4)
plot(bb_abund.2012)

@@

```{r}
bb_abund.2012.pred <- predict(bb_abund.2012)

ggplot(bb_abund.2012.pred, aes(yday, elev.mean, fit)) +
  geom_raster(aes(fill = exp(fit))) +
  #geom_contour(aes(z = fit), color = "black") +
  geom_point(data = filter(fl_abund, year == 2012), 
             aes(yday, elev.mean, size = fl.abund),
             color = "white", alpha = 0.25) +
  scale_fill_viridis_c()
```

>>>>

### FL and BB abundance / diversity
```{r echo=FALSE, include=FALSE}
### Per-transect BB abundance: species-level
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  group_by(site, date, bb.sp) %>% # group by bb.sp*transect (and other variables we want to retain)
  summarize(bb.abund = n()) %>% # get bb.sp*transect abundance
  ungroup %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) %>% # complete implicit absences
  semi_join(net, by = c("site", "date")) %>%
  mutate(year = factor(year(date))) %>%
  left_join(site_data) %>%
  mutate(yday = yday(date),
         bb.sp = factor(bb.sp),
         management = factor(management),
         site = factor(site)) %>%
  group_by(year, bb.sp) %>%
  mutate(bb.total.abund = sum(bb.abund))
```

GAM modeling of abundance per bb species
```{r}
### Pooled years
bb_abund.tw <- bam(bb.abund ~ s(bb.sp, bs = "re") + s(management, bs = "re") + year +
                     s(site, bs = "re") +
                     te(yday, elev.mean, 
                        by = bb.sp,
                        bs= c("tp", "tp"), 
                        k=c(10, 10), m=1),
                   data = bb_abund_sp,
                   discrete = TRUE,
                   family = "tw",
                   select = TRUE,
                   method = "fREML") %>% getViz()

check.gamViz(bb_abund.tw)
bb_abund.tw.sum <- summary(bb_abund.tw)
print(plot(bb_abund.tw, allTerms = TRUE), pages = 1)

### 2010
bb_abund.2010.tw <- bam(bb.abund ~ s(bb.sp, bs = "re") + s(management, bs = "re") +
                       s(site, bs = "re") +
                       te(yday, elev.mean, 
                          by = bb.sp,
                          bs= c("tp", "tp"), 
                          k=c(10, 10), m=1),
                        data = filter(bb_abund_sp, year == 2010 & bb.total.abund >= 5),
                        discrete = TRUE,
                        family = "tw",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(bb_abund.2010.tw)
bb_abund.2010.sum.tw <- summary(bb_abund.2010.tw)
print(plot(bb_abund.2010.tw, allTerms = TRUE), pages = 1)

### 2011
bb_abund.2011.tw <- bam(bb.abund ~ s(bb.sp, bs = "re") + s(management, bs = "re") +
                       s(site, bs = "re") +
                       te(yday, elev.mean, 
                          by = bb.sp,
                          bs= c("tp", "tp"), 
                          k=c(10, 10), m=1),
                        data = filter(bb_abund_sp, year == 2011 & bb.total.abund >= 5),
                        discrete = TRUE,
                        family = "tw",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(bb_abund.2011.tw)
bb_abund.2011.sum.tw <- summary(bb_abund.2011.tw)
print(plot(bb_abund.2011.tw, allTerms = TRUE), pages = 1)

### 2012
bb_abund.2012.tw <- bam(bb.abund ~ s(bb.sp, bs = "re") + s(management, bs = "re") +
                       s(site, bs = "re") +
                       te(yday, elev.mean, 
                          by = bb.sp,
                          bs= c("tp", "tp"), 
                          k=c(10, 10), m=1),
                        data = filter(bb_abund_sp, year == 2012 & bb.total.abund >= 5),
                        discrete = TRUE,
                        family = "tw",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(bb_abund.2012.tw)
bb_abund.2012.sum.tw <- summary(bb_abund.2012.tw)
print(plot(bb_abund.2012.tw, allTerms = TRUE), pages = 1)

@@

### 2020-07-13 analyses/FL_abund_GAM.Rmd
update repo

### FL and BB abundance / diversity
```{r echo=FALSE, include=FALSE}
### Per-transect BB abundance: species-level
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  semi_join(shared_transects) %>% # I only want BB data with corresponding floral data
  group_by(site, date, bb.sp) %>% # group by bb.sp*transect (and other variables we want to retain)
  summarize(bb.abund = n()) %>% # get bb.sp*transect abundance
  ungroup %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) %>% # complete implicit absences
  mutate(year = factor(year(date)))

### Per-transect floral abund: by species
fl_abund_sp <- survey %>%
  semi_join(shared_transects) %>%
  group_by(site, date, plant.sp) %>% # group by plant.sp*transect
  summarize(fl.abund = sum(flower.cover)) %>% # get plant.sp*transect total flower cover; this step really shouldn't be required, but in a few cases, there is more than one abundance entry for the same species on a given transect; not sure how that happened, but it's in the original data that Fabrice sent me; alternatively, I could drop the dups or average them. We'll see what Kalli and Fabrice have to say.
  group_by(site, date) %>%
  mutate(prop.fl.abund = fl.abund/sum(fl.abund)) %>% # get the proportional flower cover for each species.
  ungroup() %>%
  full_join(dplyr::select(net, site, date, plant.sp), by = c("site", "date", "plant.sp")) %>% # add the handful of species occurring in the visitation network but missing from the survey data
  complete(site, date, plant.sp, fill = list(fl.abund = 0, prop.fl.abund = 0)) %>%
  unique() %>%
  mutate(year = factor(year(date)))
```

@@

# All possible plants * site * date
possible_plants <- bind_rows(dplyr::select(net, c(site, date, plant.sp)),
                             dplyr::select(survey, c(site, date, plant.sp))) %>%
  unique() %>%
  complete(site, date, plant.sp)

# All possible bumbles * site * date
possible_bb <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  unique() %>%
  complete(site, date, bb.sp)

@@

fl_abund_wt <- fl_abund_sp %>%
  right_join(preference) %>% 
  mutate(fl.abund.wt = fl.abund*preference) %>%
  mutate(yday = yday(date)) %>%
  group_by(site, date, yday, bb.sp) %>%
  summarize(fl.abund.wt = sum(fl.abund.wt, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(bb_abund_sp) %>%
  left_join(bb_abund_sp_sum) %>%
  filter(bb.total.abund >= 5) %>%
  left_join(site_data, by = "site") %>%
  semi_join(survey, by = c("site", "date")) %>% # This gets rid of site*date combos that don't exist
  mutate(year = factor(year(date)), # add in the years missing from the bb_abund join
         bb.sp = factor(bb.sp), # convert bb.sp to a factor
         management = factor(management)) # convert management to a factor

@@

GAM modeling of preference-weighted abundance per bb species
## Against any potential objections to my use of an added constant to avoid the log(0) problem, I would argue that when what is being measured is a resource, the presence of a resource in vanishingly small amounts is biologically equivalent to its absence. I suppose the strength of this argument hinges on how much we trust that our samples represent our sites. But if they don't, what are we even doing here? note that this argument does not hold when one is concerned with diversity and species distributions, since then a species rarity is a very different thing that its absence.
```{r}
### Set constants
k <- min(
  filter(fl_abund_wt, fl.abund.wt > 0)$fl.abund.wt
)

### Call models
weighted_fl_abundance.2010.qp <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(fl_abund_wt, year == 2010),
                        discrete = TRUE,
                        family = "quasipoisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2010.qp)
summary(weighted_fl_abundance.2010.qp)
print(plot(weighted_fl_abundance.2010.qp , allTerms = TRUE), pages = 4)

weighted_fl_abundance.2010.qp2 <- bam(sqrt(fl.abund.wt) ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(fl_abund_wt, year == 2010),
                        discrete = TRUE,
                        family = "quasipoisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2010.qp2)
summary(weighted_fl_abundance.2010.qp2)
print(plot(weighted_fl_abundance.2010.qp2, allTerms = TRUE), pages = 3)

weighted_fl_abundance.2011.qp <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(fl_abund_wt, year == 2011),
                        discrete = TRUE,
                        family = "quasipoisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2011.qp)
summary(weighted_fl_abundance.2011.qp)
print(plot(weighted_fl_abundance.2011.qp , allTerms = TRUE), pages = 4)

weighted_fl_abundance.2011.qp2 <- bam(sqrt(fl.abund.wt) ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(fl_abund_wt, year == 2011),
                        discrete = TRUE,
                        family = "quasipoisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2011.qp2)
summary(weighted_fl_abundance.2011.qp2)
print(plot(weighted_fl_abundance.2011.qp2, allTerms = TRUE), pages = 4)

weighted_fl_abundance.2012.qp <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + management + slope.est + te(yday, elev.mean, 
                                                            by = bb.sp,
                                                            bs= c("tp", "tp"), 
                                                            k=c(10, 10), m=1),
                        data = filter(fl_abund_wt, year == 2012),
                        discrete = TRUE,
                        family = "quasipoisson",
                        select = TRUE,
                        method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2012.qp)
summary(weighted_fl_abundance.2012.qp)
print(plot(weighted_fl_abundance.2012.qp , allTerms = TRUE), pages = 4)

```

>>>>

### FL and BB abundance / diversity
```{r echo=FALSE, include=FALSE}
### Per-transect BB abundance: species-level
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  semi_join(shared_transects) %>% # I only want BB data with corresponding floral data
  group_by(site, date, bb.sp) %>% # group by bb.sp*transect (and other variables we want to retain)
  summarize(bb.abund = n()) %>% # get bb.sp*transect abundance
  ungroup %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) %>% # complete implicit absences
  mutate(year = factor(year(date))) %>%
  semi_join(net, by = c("site", "date")) # only include site-dates that actually occurred

### Per-transect floral abund: by species
fl_abund_sp <- survey %>%
  semi_join(shared_transects) %>%
  group_by(site, date, plant.sp) %>% # group by plant.sp*transect
  summarize(fl.abund = sum(flower.cover)) %>% # get plant.sp*transect total flower cover; this step really shouldn't be required, but in a few cases, there is more than one abundance entry for the same species on a given transect; not sure how that happened, but it's in the original data that Fabrice sent me; alternatively, I could drop the dups or average them. We'll see what Kalli and Fabrice have to say.
  group_by(site, date) %>%
  mutate(prop.fl.abund = fl.abund/sum(fl.abund)) %>% # get the proportional flower cover for each species.
  ungroup() %>%
  full_join(dplyr::select(net, site, date, plant.sp), by = c("site", "date", "plant.sp")) %>% # add the handful of species occurring in the visitation network but missing from the survey data
  complete(site, date, plant.sp, fill = list(fl.abund = 0, prop.fl.abund = 0)) %>%
  semi_join(survey, by = c("site", "date")) %>%  # only include site-dates that actually occurred
  unique() %>%
  mutate(year = factor(year(date)))
```

@@

# All possible plants * site * date
possible_plants <- bind_rows(dplyr::select(net, c(site, date, plant.sp)),
                             dplyr::select(survey, c(site, date, plant.sp))) %>%
  unique() %>%
  complete(site, date, plant.sp) %>%
  semi_join(survey, by = c("site", "date"))

# All possible bumbles * site * date
possible_bb <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  unique() %>%
  complete(site, date, bb.sp) %>%
  semi_join(net, by = c("site", "date"))

@@

fl_abund_wt <- fl_abund_sp %>%
  right_join(preference) %>% 
  mutate(fl.abund.wt = fl.abund*preference) %>%
  mutate(yday = yday(date)) %>%
  group_by(site, date, yday, bb.sp) %>%
  summarize(fl.abund.wt = sum(fl.abund.wt, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(bb_abund_sp) %>%
  left_join(bb_abund_sp_sum)%>%
  filter(bb.total.abund >= 5) %>% # It's not worth trying to model really rare species
  semi_join(survey, by = c("site", "date")) %>% # This gets rid of site*date combos that don't exist
  left_join(site_data, by = "site") %>%
  mutate(year = factor(year(date)), # add in the years missing from the bb_abund join
         bb.sp = factor(bb.sp), # convert bb.sp to a factor
         management = factor(management),
         site = factor(site)) %>%
  group_by(year) %>%
  mutate(total.transects.year = n()) %>%
  ungroup()

@@

GAM modeling of preference-weighted abundance per bb species
## Against any potential objections to my use of an added constant to avoid the log(0) problem, I would argue that when what is being measured is a resource, the presence of a resource in vanishingly small amounts is biologically equivalent to its absence. I suppose the strength of this argument hinges on how much we trust that our samples represent our sites. But if they don't, what are we even doing here? note that this argument does not hold when one is concerned with diversity and species distributions, since then a species rarity is a very different thing that its absence.
```{r}
### Pooled years
weighted_fl_abundance.tw <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + s(management, bs = "re") + year + 
                                  s(site, bs = "re") +
                                  te(yday, elev.mean, 
                                     by = bb.sp,
                                     bs= c("tp", "tp"), 
                                     k=c(10, 10), m=1),
                                data = fl_abund_wt,
                                discrete = TRUE,
                                family = "tw",
                                select = TRUE,
                                method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.tw)
weighted_fl_abundance.tw.sum <- summary(weighted_fl_abundance.tw)
print(plot(weighted_fl_abundance.tw, allTerms = TRUE), pages = 1)
saveRDS(weighted_fl_abundance.tw, "../analyses/output/weighted_fl_abundance.rds")

### 2010
weighted_fl_abundance.2010.tw <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + s(management, bs = "re") + 
                                       s(site, bs = "re") +
                                       te(yday, elev.mean, 
                                          by = bb.sp,
                                          bs= c("tp", "tp"), 
                                          k=c(10, 10), m=1),
                                     data = filter(fl_abund_wt, year == 2010),
                                     discrete = TRUE,
                                     family = "tw",
                                     select = TRUE,
                                     method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2010.tw)
weighted_fl_abundance.2010.tw.sum <- summary(weighted_fl_abundance.2010.tw)
print(plot(weighted_fl_abundance.2010.tw , allTerms = TRUE), pages = 1)

### 2011
weighted_fl_abundance.2011.qp <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + s(management, bs = "re") + 
                                       s(site, bs = "re") +
                                       te(yday, elev.mean, 
                                          by = bb.sp,
                                          bs= c("tp", "tp"), 
                                          k=c(10, 10), m=1),
                                     data = filter(fl_abund_wt, year == 2011),
                                     discrete = TRUE,
                                     family = "quasipoisson",
                                     select = TRUE,
                                     method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2011.qp)
weighted_fl_abundance.2011.qp.sum <- summary(weighted_fl_abundance.2011.qp)
print(plot(weighted_fl_abundance.2011.qp , allTerms = TRUE), pages = 1)

### 2012
weighted_fl_abundance.2012.qp <- bam(fl.abund.wt ~ s(bb.sp, bs = "re") + s(management, bs = "re") + 
                                       s(site, bs = "re") +
                                       #slope.est + 
                                       te(yday, elev.mean, 
                                          by = bb.sp,
                                          bs= c("tp", "tp"), 
                                          k=c(10, 10), m=1),
                                     data = filter(fl_abund_wt, year == 2012),
                                     discrete = TRUE,
                                     family = "quasipoisson",
                                     select = TRUE,
                                     method = "fREML") %>% getViz()

check.gamViz(weighted_fl_abundance.2012.qp)
weighted_fl_abundance.2012.qp.sum <- summary(weighted_fl_abundance.2012.qp)
print(plot(weighted_fl_abundance.2012.qp , allTerms = TRUE), pages = 1)
```
